%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

A first question we might ask when studying algebraic varieties is its dimension. 

In the case of secant varieties, we can easily define a notion of expected dimension by a simple parameter count. We will see that this is not always the actual dimension: examples of varieties with secant varieties of dimension lower than the expected are classically known since the XIX century. We will explain how the problem of classifying such special varieties has been approached.

Recall that, knowing dimensions of secant varieties of a non-degenerate algebraic variety $X$ allows us to deduce the generic $X$-rank, see \ref{geometrySecants-section-rank} and in particular \ref{geometrySecants-definition-generic_rank}. This is one reason why the problem of classifying defective Veronese, Segre, Segre-Veronese varieties and Grassmannians received a particular attention. 

\section{Expected Dimension}
\label{geometrySecants-section-expectedDimension}
% Author: Alessandro Oneto

\begin{lemma}
\label{geometrySecants-lemma-expecteddimension}
% Author: Alessandro Oneto
Let $X_1\vvirg X_s \subseteq \bbP^N$ be algebraic varieties. Then 
\[
    \dim j(X_1\vvirg X_s) \leq \min\{ N, \smallsum_{i=1}^s \dim(X_i) + s-1\}.
\]
In particular, for any algebraic variety $X \subseteq \bbP^N$,  
\[
    \dim \sigma_s(X) \leq \min\{ N , s\dim(X) + s - 1\}.
\]
\end{lemma}
\begin{proof}
As they can be obtained by projection, the join $j(X_1,\ldots,X_s)$ and the secant variety $\sigma_s(X)$ have dimension smaller than their abstract analogous.

The dimension of the abstract join $\itJoin(X_1,\ldots,X_s)$, as defined in \ref{geometrySecants-section-secants}, easily follows from the fact that locally it can be interpreted as a $\bbP^{s-1}$-bundle over $X_1 \ttimes X_s$: therefore, 
    \[
        \dim \itJoin(X_1,\ldots,X_s) = \sum_{i=1}^s \dim(X_i) + s-1.
    \]
    In particular, $\dim \Sec_s(X) = s\dim(X) + s-1$.
\end{proof}

One may {\it expect} that the upper bound of \ref{geometrySecants-lemma-expecteddimension} is the actual dimension. This leads the the following definition. 
\begin{definition}
\label{geometrySecants-definition-expecteddimension}
% Author: Alessandro Oneto
    Let $X_1\vvirg X_s \subseteq \bbP^N$ be algebraic varieties. The {\it virtual dimension} of $j(X_1,\ldots,X_s)$ is the dimension of the abstract join. Its {\it expected dimension} is the minimum between the virtual dimension and the dimension of the ambient space. In other words,
    \begin{align*}
        \virdim j(X_1,\ldots,X_s) & = \sum_{i=1}^s \dim(X_i) + s-1; \\ 
        \expdim j(X_1,\ldots,X_s) & = \min \{N,\virdim j(X_1,\ldots,X_s)\}.
    \end{align*}
    In particular, if $X \subseteq \bbP^N$ is an algebraic variety, then
    \begin{align*}
        \virdim \sigma_s(X) & = s\dim(X) + s-1; \\ 
        \expdim \sigma_s(X) & = \min \{N,\virdim \sigma_r(X)\}.
    \end{align*}
By \ref{geometrySecants-lemma-expecteddimension} the expected dimension is always an upper bound for the dimension of a secant variety. We say that $X$ is \emph{$s$-defective} if $\dim \sigma_s(X) < \expdim\sigma_s(X)$. We call the difference $\delta_s(X) = \expdim\sigma_s(X) - \dim \sigma_s(X)$ the \emph{$s$-defect} of $X$. 

%% We will also give an algebraic meaning for the difference $\vdim\sigma_s(X) - \dim \sigma_s(X)$ when we will look at dimensions of secant varieties in terms of interpolation problems. 
\end{definition}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{First lemmas and non-defectiveness of curves}
\label{geometrySecants-section-first_properties}
% Author: Alessandro Oneto
We collect here the first properties on secant varieties and their dimensions. The following lemmas were all classically know, see \cite{Pal09}. For a recent reference, we refer to \cite{Rus16}.

First of all, we observe that linear spaces are the \emph{idempotent} varieties in terms of secant varieties. 

\begin{lemma}
    \label{geometrySecants-lemma-palatini_1}
% Author: Alessandro Oneto
    Let $X\subset \bbP^N$ be an irreducible projective algebraic variety. If $\sigma_{r+1}(X) = \sigma_r(X)$, then $\sigma_r(X)$ is a projective linear space. 
\end{lemma}
\begin{proof}
    We prove this in three steps.

    {\bf Claim 1.} The statement is true for $r = 1$. I.e., if $\sigma_2(X) = X$ then $X$ is a projective linear space.
    \begin{proof}[Proof of Claim 1.]
        Let $P \in \langle X \rangle$. Then, $P = \lambda_1P_1 + \ldots + \lambda_rP_r$ for some $P_i \in X$ and $\lambda_i \in \Bbbk$. By assumption, $\lambda_1P_1 + \lambda_2P_2 \in X$ and then, $(\lambda_1P_1 + \lambda_2P_2) + \lambda_3P_3 \in X$. Inductively, we deduce that $P \in X$. Hence, by generality of $P$, $\langle X \rangle \subset X$ and equality follows. 
    \end{proof}

    {\bf Claim 2.} If $\sigma_{r+1}(X) = \sigma_r(X)$ then $\sigma_{r+h}(X) = \sigma_r(X)$ for all $h \geq 1$. 
    \begin{proof}[Proof of Claim 2.]
        Let $P \in \sigma_{r+2}(X)$, then $P = \lambda_1P_1 + \ldots + \lambda_{r+1}P_{r+1} + \lambda_{r+2}P_{r+2}$ where $P_i \in X$. By assumption, $\lambda_1P_1 + \ldots + \lambda_{r+1}P_{r+1} = \mu_1Q_1 + \ldots + \mu_rQ_r$, where $Q_i \in X$. In particular, $P \in \sigma_{r+1}(X)$. Therefore, $\sigma_{r+2}(X) \subset \sigma_{r+1}(X)$ and equality follows. 
    \end{proof}

    We are now ready to prove the general statement. By Claim 2, we have that $\sigma_{2r}(X) = \sigma_r(X)$. Since $\sigma_2(\sigma_r(X)) \subset \sigma_r(X)$, by irreducibility, we deduce that $\sigma_2(\sigma_r(X)) = \sigma_r(X)$. By Claim 1, $\sigma_r(X)$ is a projective linear space. 
\end{proof}


\begin{lemma}
    \label{geometrySecants-lemma-palatini_2}
% Author: Alessandro Oneto
Let $X\subset \bbP^N$ be a irreducible projective variety. If $\dim\sigma_{r+1}(X) = \dim\sigma_r(X)+1$, then $\sigma_{r+1}(X)$ is a projective linear space. 
\end{lemma}
\begin{proof}
    Let $p \in X$ be a general point. Hence,
    \[
        \sigma_s(X) \subsetneq j(p, \sigma_s(X)) \subset \sigma_{s+1}(X).  
    \]
    By assumption and irreducibility of $\sigma_{s+1}(X)$, $j(p, \sigma_s(X)) = \sigma_{s+1}(X)$. Let $q \in \sigma_{s+1}(X)$ be a generic smooth point: by the latter equality, $q \in \langle p,z \rangle$ for some $z \in \sigma_s(X)$. Being a cone, we also get that $p \in T_q\sigma_{s+1}(X)$. By generality of $p \in X$, we conclude that $\langle X \rangle \subset \sigma_{s+1}(X)$ and then, since the opposite inclusion is trivial, $\sigma_{s+1}(X) = \langle X \rangle$ and $\sigma_s(X)$ is a hypersurface. 
\end{proof}


An immediate but meaningful corollary is the fact that, if $X$ is \emph{non-degenerate}, i.e., it is not contained in any proper linear subspace, then the secant varieties of $X$ eventually fill the ambient space. 

\begin{lemma}
    \label{geometrySecants-lemma-secants_of_non_degenerate}
% Author: Alessandro Oneto
    Let $X\subset \bbP^N$ be a non-degenerate algebraic variety. Then, 
    \[
        X \subsetneq \sigma_2(X) \subsetneq \cdots \subsetneq \sigma_r(X) = \bbP^N.
    \]
\end{lemma}
As already mentioned, this observation allows us to conclude well-define the generic rank with respect to non-degenerate algebraic varieties, see \ref{geometrySecants-definition-generic_rank}.

Another important corollary is that \emph{curves are never defective.}
\begin{proposition}
    \label{geometrySecants-proposition-curves_nondefectiveness}
% Author: Alessandro Oneto
    Let $X\subset \bbP^N$ be a projective curve. Then, $\dim \sigma_s(X) = \min\{N, 2s-1\}$.
\end{proposition}
\begin{proof}
    By \ref{geometrySecants-lemma-expecteddimension},
    \[
        \dim \sigma_s(X) \leq s\dim(X) + s - 1 = 2s - 1.
    \]
    By \ref{geometrySecants-lemma-secants_of_non_degenerate} and the non-degeneracy assumption,
    \[
        \dim \sigma_s(X) \geq \dim \sigma_{s-1}(X) + 2
    \]
    unless $\sigma_{s-1}(X)$ is a hypersurface. Therefore, as long as $\sigma_{s-1}(X)$ is not an hypersurface, we proceed by induction (the base step $\sigma_1(X) = X$ being trivial) to obtain
    \[
        2s-1 \geq \dim \sigma_s(X) \geq \dim \sigma_{s-1}(X) + 2 = 2(s-1) - 1 + 2 = 2s - 1
    \]
    which implies $\dim \sigma_s(X) = 2s-1$. If $\sigma_{s-1}(X)$ is an hypersurface, then we clearly have that $\dim \sigma_s(X) = N = \min\{N,2s-1\}$ where the latter equality follows from being $2(s-1)-1 = 2s-3 = N - 1$. 
\end{proof}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Terracini's Lemma}
\label{geometrySecants-section-Terracini}
% Author: Alessandro Oneto

A general approach to study dimensions of secant varieties is through their tangent spaces at general points. For this reason, a fundamental tool in the study of dimension of secant varieties is \emph{Terracini's Lemma}, dating back to \cite{Ter11}.

The following statement is a modern approach which can be found in \cite[Corollary 1.10]{Adl87} or in the books \cite[Proposition II.1.10]{Zak93} and \cite[Theorem 1.4.1]{Rus16}.

\begin{lemma}[Terracini's Lemma]
\label{geometrySecants-lemma-terracini}
% Author: Alessandro Oneto
    Let $X, Y \subseteq \mathbb{P}^N$ be irreducible projective varieties over a field $\Bbbk$ of characteristic $0$. Let $x \in X, y \in Y$ be generic points and let $z \in \langle x,y \rangle$ be a generic point. In particular, we may assume that $z$ is a smooth point of $j(X,Y)$. Then
    \[
        T_{z} j(X,Y) = \langle T_x X , T_y Y\rangle.
    \]
\end{lemma}
\begin{proof}
We record the proof over the field of complex numbers, which is the setting originally studied by Terracini and which provides a good intuition on why the result works. See also \cite{BO08}. 

Consider the local parametrization of $j(X,Y)$ at $z$ given by the corresponding linear combination of the parametrization of $X$ and $Y$ at $x$ and $y$, respectively. The statement follows by observing that the matrix obtain by partial derivatives and whose rows span the tangent space of $j(X,Y)$ at $z$ is obtained by elementary operation on the concatenation of the corresponding matrices for $X$ and $Y$. 
\end{proof}

\begin{example}
\label{geometrySecants-example-terracini_positive_char}
% Author: Alessandro Oneto
    In positive characteristics, only the inclusion $T_{z} j(X,Y) \supset \langle T_x X , T_y Y\rangle$ is guaranteed, see e.g. \cite[Proposition II.1.12]{Zak93}. For example, over $\Bbbk$ with ${\rm char}(\Bbbk)= p > 0$ consider the curve $C \subset \bbP^3$ which is the projective closure of the curve parametrized by $t \mapsto (t,t^p,t^{p^2}) \subset \bbA^3$. Note that all tangent lines are parallel, i.e., there exists a point lying on all tangent lines to the curve $C$: such a curve is called {\it strange} in Section IV.3 of \cite{Har77}. It follows that the linear span on two tangent lines is $2$-dimensional since they have non-trivial intersection, while the secant variety of $C$, and then a general tangent space to it, is $3$-dimensional by Proposition \ref{geometrySecants-proposition-curves_nondefectiveness}. This contradicts the statement of Terracini's Lemma (Lemma \ref{geometrySecants-lemma-terracini}).
\end{example}

\begin{remark}
\label{geometrySecants-remark-terracini}
% Author: Alessandro Oneto
    Terracini's Lemma tells us that the dimension of $j(X,Y)$ is as in \ref{geometrySecants-definition-expecteddimension} if and only if two general tangent spaces $T_x(X)$ and $T_y(Y)$ do not intersect, unless $j(X,Y)$ fills the ambient space. 
\end{remark}

In the case $X = Y$, Terracini's Lemma can be used to study dimensions of secant varieties. As a first example, we can give a first explanation why secant varieties of quadratic Veronese varieties are always defective. In order to do that, we first need to understand what is the tangent space to Veronese varieties. 

\begin{example}[Quadratic Veronese varieties are always defective]
\label{geometrySecants-lemma-quadratic_Veronese_defective}
% Author: Alessandro Oneto
    Let $V$ be a complex linear space. Let $\nu_d(\bbP V) \subset \bbP(S^d V)$ be the degree-$d$ Veronese embedding of $\bbP^n$ as in \ref{introduction-definition-Veronese}. Let $v \in V$. Then
    \[
        T_{[v]}\nu_d(\bbP V) = \{[v^{d-1}v'] ~:~ v' \in V\}. 
    \]
    Indeed, it is enough to observe that, for any $v' \in V$,
    \[
        \left.\frac{d}{dt}\right|_{t = 0}(v+tv')^{d} = dv^{d-1}v'.
    \]
    We consider now the quadratic case $d = 2$ and we employ Terracini's Lemma. Let $[v_1],\ldots,[v_s] \in \nu_2(\bbP V)$ be general points. We immediately note that 
    \[
        T_{[v_i]}\nu_2(\bbP V) \cap T_{[v_j]}\nu_2(\bbP V) = [v_iv_j]
    \]
    while, by dimension, we would expect to have empty intersection (if $\dim V \geq 3$). Therefore, we immediately deduce that quadratic Veronese varieties are defective, in particular:
    \[
        \dim \sigma_r(\nu_2(\bbP V)) \leq \min \{\dim S^dV - 1, r\dim \bbP V + r - 1 - {r \choose 2}\}.  
    \]
\end{example}
