Tensors manifest in various forms across mathematics and numerous other fields. The roots of the study of tensors and their properties goes back by more than a century: for example, in 1850s, Sylvester was already considering {\it decompositions of binary forms as sums of powers} \cite{Syl51}; or, since the late XIX century, classical algebraic geometers were already studying {\it secant varieties} \cite{Sco08,Ter11}; or, in 1920s, Hitchcock already introduced {\it polyadic decompositions} of tensors \cite{Hit27}.

Considering the long story spanning multiple domains, there is are many beautiful and rich resources available for both beginners and experts and written in different languages with varying jargon. In particular, in recent years, we have witnessed a remarkable growth in literature pertaining to what can broadly be called \emph{geometry of tensors}. Scholars from diverse fields such as classical geometry, theoretical computer science, data science, and quantum physics gave their contribution, actively engaging in each other's open problems and questions. However, due to the varied languages used in these works, gaining a comprehensive overview of the field's (or better, the fields') current state presents a significant challenge. Just to mention some of them, we refer to \cite{Zak93,Ger96,IK99,Hac12,Lan12,CGO14,Rus16,BCCGO18,BBCMM19}. 

It was from this realization that the concept of a unifying portal was born. The \textbf{MultilinearVerse} wants to be an online platform conceived with the aim of serving as such a unifying resource with the possibility of growing in the future to cover all different facets of the subject and dynamically follow the advancement of state-of-the-art research.

The development of this portal started in early 2024, drawing considerable inspiration from The Stacks project \cite{Stacks}. The portal is built on the Gerby project, an online tag-based view for large \LaTeX~documents. The tag-based system ensures that every page, statement, and \emph{label} within the document is assigned its own unique tag, facilitating easy referencing. MultilinearVerse is not a book and it does not intend to replace classical resources. It has chapter, sections, and a linear numbering of its statements. Yet, its design encourages a tree-based approach to navigation, based on the mathematical connections between different topics, which are made readily apparent through various interlinks.

We do not plan that the portal will cover background material, to the level of a Master's degree in Mathematics. We will include more precise references on the background in the single sections of the portal. We acknowledge that the definition of \emph{background material} will be mostly arbitrary. After all, MultilinearVerse is a product of its developers, and necessarily reflects their mathematical strengths and weaknesses. Most of the geometry is limited to the complex setting, with a focus on results and methods coming from classical algebraic geometry. We expect and we hope that \emph{all} aspects of the multifaceted world of tensors will be covered eventually with the contribution of our rich community. 

\chapter{Preliminaries}
\label{introduction-chapter-intropreliminaries}
\input{preliminaries.tex}
